---
title: "02_data_manipulation"
output: html_document
date: "2026-09-24"
---

```{r}
library(tidyverse)
```

```{r}
litters_df =
  read_csv(file = "data/FAS_litters.csv", 
           na = c("", "NA", ".")
  )
litters_df = janitor::clean_names(litters_df)

pups_df = 
    read_csv("data/FAS_pups.csv",
             na = c("", "NA", "."),
             skip = 3
    )
pups_df = janitor::clean_names(pups_df)
```

```{r}
select(litters_df, group, litter_number)

select(litters_df, group:gd_of_birth)

select(litters_df, -group)

select(litters_df, starts_with("gd"))

select(litters_df, starts_with("gd"), group)

select(litters_df, GROUP = group, everything())

select(litters_df, - starts_with("gd"))
```

a couple of alternatives
```{r}
rename(litters_df, GROUP = group)

relocate(litters_df, litter_number, gd0_weight)
#putting columns in a new order
```

in the pups dataset, select litter number, sex, and PD ears
```{r}
select(pups_df, litter_number, sex, pd_ears)
```

## 'filter' some rows
```{r}
filter(litters_df, gd_of_birth == 19)

filter(litters_df, pups_born_alive <= 5)

filter(litters_df, pups_born_alive != 6)

filter(litters_df, group %in% c("Con7", "Con8", pups_born_alive <7))

drop_na(litters_df, gd0_weight)
```


in the pups datsaset, keep only rows wehre PD walk is less than 11 and sex is 2
```{r}
filter(pups_df, pd_walk < 11, sex == 2)
```

## 'mutate' to change some columns

add and change some columns
```{r}
mutate(
  litters_df, 
  wt_gain = gd18_weight - gd0_weight,
  group = str_to_lower(group))
```

in the pups dataset, create a vairable that subtracts 7 from PD walk
```{r}
mutate(pups_df, pd_walk_minus_7 = pd_walk -7)
```

## 'arrange' the data

```{r}
arrange(litters_df, pups_born_alive)
arrange(litters_df, desc(pups_born_alive))
```

```{r}
## shift command m

litters_df = 
  read_csv("data/FAS_litters.csv", na = c("","NA",".")) |> 
  janitor::clean_names() |> 
  select(group, starts_with("gd")) |> 
  drop_na() |> 
  mutate(
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group)
  )
```

load pups, clean names, drop missing, keep litter number and pd variables, add pd walk -7

pups_df = 
  read_csv("data/FAS_pups.csv", skip = 3, na = c("","NA",".")) |> 
  janitor::clean_names |> 
  drop_na |> 
  select(litter_number, starts_with("pd")) |> 
  mutate(
  pd_walk = pd_walk - 7
  )


```{r}
litters_df |> 
  lm(gd18_weight ~ gd0_weight, data = _)
```

```{r}
litters_df |> 
  select(group)

litters_df |> 
  pull(group)
```

never do '$'
```{r}
litters_df$group
```

pipes and viewung
```{r}
litters_df |> 
  filter(group %in% c)
```





